\chapter{Preface}
\label{preface}

\section{The Philosophy}

\para{This is a collection of notes written by me on all of the four subjects included in mainstream olympiads. 
Much of the contents is sourced from a various plethora of free and paid books.}

\parbreak

\para{This is not meant to be a one-spot go to for all your Olympiad preparation, 
even though I have tried to include every bit of information I could get my hands on. 
One important part that is missing from it are problems. 
The Primary reason for that being my laziness and the fact and these are meant to be my \emph{notes}, 
which shall capture my Olympiad journey. This is not an olympiad go-to book.
Be aware}

\section{Acknowledgments}

\para{First and foremost, my greatest thanks goes to Evan Chen for his \href{https://github.com/vEnhance/dotfiles/blob/main/texmf/tex/latex/evan/evan.sty}%
{awesome style file}, parts
of which I have modified and used in my notes. He has also been a source of much inspiration,
and for the EGMO \cite{egmo} book, upon which much of the Geometry is based.}

\para{I also must thank Aditya Khurmi for his freely available book MONT \cite{mont},
Although he did not have any particular ivolvement here, I must also thank Kevin Zhou, for 
inspiring me.}

\section{Notation}

\begin{itemize}
    \ii  \(\NN = \{1, 2, 3 \ldots \}\)
    \ii \(\NN_0 =\{0, 1, 2, \ldots\}\)
\end{itemize}

\clearpage
\thispagestyle{empty}
    \vspace*{\fill}
    {\em \centering \hspace{4.5cm} To her, whose death still haunts me.}
    \vspace*{\fill}
