\chapter{Elementary Prereqs}
\section{Intervals}

\marginpar{\raggedleft The notation \(][\) is French, and is less commonly used in India, USA, etc.}

\begin{definition}
    For \(a, b, x \in \RR \) and \(a < b\),
    \begin{enumerate}
        \ii The subset \(\{x \mid a < x < b\}\) of \(\RR\) is called the open interval of \(a, b\)
        and is represented as \((a,b)\) or \(]a,b[\).
        \ii The subset \(\{x \mid a \le x \le b \}\) of \(\RR\) is called the closed interval
        of \(a, b\) and is represented as \([a,b]\).    
    \end{enumerate}
    
\end{definition}

\todo{Add figures for all}

\begin{figure}[ht]
    \centering
    \begin{subfigure}{0.31\textwidth}
        \begin{tikzpicture}
            \begin{axis}[funcgraph]
            \end{axis} 
            \draw[latex-] (-1,0) -- (9,0) ;
            \draw[-latex] (-1,0) -- (9,0) ;
            \foreach \x in  {0,1,2,3,4,5,6,7,8} \draw[shift={(\x,0)},color=black] (0pt,3pt) -- (0pt,-3pt);
            \foreach \x in {0,1,2,3,4,5,6,7,8} \draw[shift={(\x,0)},color=black] (0pt,0pt) -- (0pt,-3pt) node[below] {$\x$};
            \draw[*-o] (2.9,0) -- (5.1,0);
            \draw[very thick] (2.98,0) -- (5,0);
        \end{tikzpicture}    
    \end{subfigure}
    
    \caption{Closed-Open Interval}
\end{figure}

\para{We can make all kind of closed-open (\(a \le x < b\)) or open-closed (\(a < x \le b\))
intervals, but there is no point in giving them a name.}

\section{Inequalities}

\para{Real numbers and all its subsets have a common property. They're ordered under the
\(<, >, =\) relations. Thus, for any two real numbers, we can always compare them and have the
trichotomy :}
\begin{claim}
    For \(a, b \in \RR\) we will always have, 
    \begin{enumerate}
        \ii \(a < b\),
        \ii \(a = b\), or
        \ii \(a > b\).
    \end{enumerate}
\end{claim}

\para{This why inequalities are so useful for reals! We can always form inequalities 
for equations. Now, let us propose the following properties of inequalities.}

\begin{proposition}
    For \(a, b \in \RR\),
    \begin{enumerate}
        \ii \(a < b \iff a\pm k < b \pm k \)
        \ii \(a < b, c < d \iff a + c < b + d\)
        \ii \(a < b, c < d \iff a - d < b - c\)
        \ii \(a < b, k > 0 \implies ka<kb\)
        \ii \(a < b, k < 0 \implies ka>kb\) 
        \ii \(a < b, b < c \implies a < c\)
        \ii \(0 < a < b, r > 0 \implies a^r < b^r\)
        \ii \(0 < a < b, r < 0 \implies a^r > b^r\)
        \ii \(a > 0 \iff a + \dfrac{1}{a} \ge 2\)
        \ii \(a < 0 \iff a + \dfrac{1}{a} \le 2\)         
    \end{enumerate}
\end{proposition}

\subsection{Inequalities concerning squares}

\para{One thing of note to consider is how do the inequalities of \(x\) from
\(x^2\) follow? Consider \( a \le x \le b\). You might expect that for \(a \le x \le b\),
\(a^2 \le x \le b^2\). But that is simply incorrect! What we instead get is that \(x^2\) lies
between \(0\) and \(\max(a^2,b^2)\). Here, max represents that \(x^2\) lies between \(0\) and
whichever of the \(a^2\), \(b^2\) is greater.}
\para{In-fact, if \(a \le x \le b \iff a^2 \le x^2 \le b^2 \), then \(a\), \(b\) and hence \(x\)
must be \(\ge 0\)!}
\parbreak
\para{Also, if \(x^2 \le a^2\), then \(x \in [-a,a]\)\\
And, if \(x^2 \ge a^2\), then \(x \in (-\infty,-a] \cup [a, \infty]\)}



\begin{example}
    For \(-5 \le x \le 4\), we have \( 0 \le x^2 \le \max(-5^2, 4^2)\). Therefore, \(0 \le 
    x^2 \le 25\).
\end{example}

\begin{figure}[ht]
    \centering
    \begin{tikzpicture}[scale=0.86]
        \begin{axis}[funcgraph]
            \addplot[color=black,domain=-6:-5]{x^2};
            \addplot[color=red,domain=-5:4]{x^2};
            \addplot[color=black,domain=4:6]{x^2};
        \end{axis}
    \end{tikzpicture}
    \caption{\(x^2\) in red for \(-5 \le x \le 4\)}
\end{figure}

